---
title: "Predicting Sepsis in the ICU"
author: Ryan Quan and Frank Chen
date: "April 30, 2015"
output: html_document
---

```{r, message=FALSE}
library(plyr)
library(dplyr)
library(reshape2)
library(stringr)
library(magrittr)
library(lubridate)
library(knitr)
library(ggplot2)
library(caret)
library(rpart)
library(e1071)
library(randomForest)
opts_knit$set(cache.extra = rand_seed)
```

# Methods

## Data Source

```{r, message=FALSE}
# devtools::add_path('~/usr/local/bin/psql')
# devtools::install("~/Downloads/rpostgresql-read-only/RPostgreSQL/")
db <- src_postgres(dbname = "MIMIC2",
                  host = "ec2-54-163-173-71.compute-1.amazonaws.com",
                  port = 5432,
                  user = "ec2-user",
                  password = "thisisalongpassword1234")
```

## Population

Since our objective is to train a prediction model that will detect the onset of sepsis within the first few hours of admittance to the ICU, we exclude patients who were not admitted to the ICU for the first sequence of their hospital visit. This is denoted by the column `icustay_first_flg`. Moreover, we restrict our analysis to the adult age group, which is defined as patients greater than or equal to 15 years of age. 

## Outcomes

To define positive cases of sepsis, we lumped ICD-9 codes for `sepsis`, `severe sepsis`, and `septic shock`, as these are all part of the "sepsis continuum". 

```{r caseid, cache=TRUE}
query <- sql(
"SELECT subject_id, hadm_id, icustay_id
FROM mimic2v26.icustay_detail
WHERE (icustay_first_flg = 'Y' AND subject_icustay_seq = '1' AND icustay_age_group = 'adult')
AND subject_id IN
    (SELECT DISTINCT subject_id
    FROM mimic2v26.icd9
    WHERE (code LIKE '995.9%' OR code = '785.52'))"
)
caseID <- tbl(db, query) %>%
    collect() %>%
    mutate(label = "case")
```

For our controls, we randomly sampled 5000 patients from the MIMIC II database who meet the same exclusionary criteria.

```{r controlid, cache=TRUE}
query <- sql(
"SELECT subject_id, hadm_id, icustay_id
FROM mimic2v26.icustay_detail
WHERE (icustay_first_flg = 'Y' AND subject_icustay_seq = '1' AND icustay_age_group = 'adult')"
)
controlID <- tbl(db, query) %>%
    filter(!(subject_id %in% caseID$subject_id)) %>%
    collect() %>%
    mutate(label = "control")
```

With this, we now create a data frame that contains the labels and unique identifiers for our training set.

```{r, cache=TRUE}
set.seed(1)
randomSubset <- sample(controlID$subject_id, 5000)
controls <- controlID[controlID$subject_id %in% randomSubset, ]
ids <- rbind(caseID, controls)
```

## Confounding Medical Interventions

Since our objective is to predict the onset of sepsis, we must also exclude patients who received antibiotics upon admission into the ICU. If a patient received any antibiotics within the first 24 hours, he or she was dropped from the training set. Doing so will allow us to avoid bias introduced via confounding medical interventions.

```{r microbiology, cache=TRUE}
query <- sql(
"SELECT a.subject_id AS subject_id, a.hadm_id AS hadm_id, a.ab_itemid AS ab_itemid, 
        a.charttime - b.admit_dt AS sepsis_flag
FROM mimic2v26.microbiologyevents AS a
JOIN mimic2v26.admissions AS b ON a.subject_id = b.subject_id
AND a.hadm_id = b.hadm_id"
)
microbiology <- tbl(db, query) %>%
    filter(hadm_id %in% ids$hadm_id) %>%
    filter(sepsis_flag < "1 day") %>%
    group_by(subject_id, hadm_id) %>%
    summarise(sepsis_flag = n()) %>%
    collect()
```

We now update our labels to reflect this exclusion. As you can see, our `cases` drop substantially. We will need to adjust this class imbalance somehow.

```{r}
ids %<>% filter(!(hadm_id %in% microbiology$hadm_id))
table(ids$label)
```

## Feature Extraction

Since we are interested in creating a classifier that can predict the onset of sepsis using accessible and available clinical data, we restrict our feature set to the `demographic`, `icustay_details`, and `chartevents` tables in the MIMIC II database.

### Demographic

We extract all variables from the `demographic` table, which includes items like religious affiliation, insurance information, and marriage status. From the `icustay_detail` table, we extract patient information that was recorded upon admission, such as age, admission time, weight, height, etc.

```{r demographic, cache=TRUE}
query <- sql(
"SELECT *
FROM mimic2v26.demographic_detail AS a
JOIN mimic2v26.icustay_detail AS b ON a.subject_id=b.subject_id
AND a.hadm_id=b.hadm_id"
)
demographic <- tbl(db, query) %>%
    filter(icustay_id %in% ids$icustay_id) %>%
    select(icustay_id, gender, icustay_intime, icustay_admit_age, contains("descr"), contains("first")) %>%
    collect()
```

```{r}
demographic %<>%
    select(-hospital_first_flg, -icustay_first_flg) %>%
    filter(icustay_admit_age < 120) %>%
    mutate(icustay_intime = ifelse(hour(icustay_intime) %in% seq(6, 11), "morning",
                            ifelse(hour(icustay_intime) %in% seq(12, 5), "afternoon",
                            ifelse(hour(icustay_intime) %in% seq(6, 11), "evening", "night")))) 

collapseRace <- function(race) {
    if (str_detect(race, "white")) return("white")
    else if (str_detect(race, "black")) return("black")
    else if (str_detect(race, "asian")) return("asian")
    else if (str_detect(race, "hispanic")) return("hispanic")
    else if (str_detect(race, "unable") | str_detect(race, "unknown")) return("unknown")
    else return("other")
}

demographic$ethnicity_descr <- sapply(tolower(demographic$ethnicity_descr), 
                                      collapseRace, 
                                      USE.NAMES=FALSE)

```



## Chart Events

For our chart events, we discretize routine clinical measurements taken over the first 24 hours for every hour. 

```{r chartevents, cache=TRUE}
query <- sql(
"SELECT subject_id, icustay_id, itemid, charttime, value1num 
FROM mimic2v26.chartevents 
WHERE itemid IN (618, 646, 677, 52, 456, 211, 20002, 1542, 677)"
)

chartevents <- tbl(db, query) %>% 
	filter(icustay_id %in% ids$icustay_id) %>%
    ## hour and minutes are lost if not coerced
	mutate(timeString = as.character(charttime)) %>% 
	select(-charttime) %>% 
	collect() %>%
    mutate(charttime = ymd_hms(timeString),
           timeString = NULL) %>%
	group_by(subject_id, icustay_id, itemid) %>%
    ## set reference time and discretize by the hour
	mutate(groundTime = min(charttime),
		   intHour = as.numeric(difftime(charttime, groundTime, units="hours")), 
		   intHourR = round(intHour)) %>%
    group_by(subject_id, icustay_id, itemid, intHourR) %>%
	summarize(value = mean(value1num)) %>%
    ungroup()
```

Routine clinical measurements along with their respective `itemid` code are shown below:

```{r}
query <- sql(
"SELECT itemid, label
FROM mimic2v26.d_chartitems
WHERE itemid IN (618, 646, 677, 52, 456, 211, 20002, 1542, 677)"
)
chartLabels <- tbl(db, query) %>%
    collect()
chartLabels
```

```{r}
## tidy up label formatting
chartLabels %<>%
    mutate(label = tolower(label),
           label = gsub(" ", "_", label),
           label = ifelse(label == "temperature_c_(calc)", "temperature", label)) %>%
    filter(!(label %in% c("respiratory_sofa_score", "arterial_bp_mean")))
charteventsDiscrete <- chartevents %>%
    join(chartLabels, by="itemid", type="inner")
```

```{r chartevents2, cache=TRUE}
summarizeByHour <- function(features, hour) {
    summarizedFeatures <- features %>%
        select(icustay_id, label, intHourR, value) %>%
        filter(intHourR >= 0 & intHourR <= hour) %>%
        group_by(icustay_id, label) %>%
        summarize(mean = mean(value, na.rm=TRUE),
                  min = min(value, na.rm=TRUE),
                  max = max(value, na.rm=TRUE),
                  std = sd(value, na.rm=TRUE)) %>%
        melt(id.vars=c("icustay_id", "label")) %>%
        dcast(icustay_id ~ label + variable, value.var="value")
    return(summarizedFeatures)
}
hours <- c(1, 3, 6, 12, 24)
featuresList <- lapply(hours, function(hour) {
    summarizeByHour(charteventsDiscrete, hour=hour)
    })
```

## Feature Matrix

```{r features}
featuresList <- lapply(featuresList, join, demographic, by="icustay_id")
featuresList <- lapply(featuresList, join, ids[, c("icustay_id", "label")])
```


```{r}
features <- featuresList[[3]]
row.names(features) <- features$icustay_id
features <- features[ , -grep("icustay_id", names(features))]
continuousVars <- names(which(sapply(features, is.numeric)))
categoricalVars <- names(features)[!(names(features) %in% continuousVars)]

set.seed(1)
inTrain <- createDataPartition(features$label, times=1, p=0.7, list=FALSE)
training <- features[inTrain, ]
testing <- features[-inTrain, ]


preProcessed <- preProcess(training[, continuousVars],
                           method=c("center", "scale", "medianImpute"),
                           na.remove=FALSE)
what <- predict(preProcessed, training[, continuousVars])   
nrow(what)

what

shit <- predict(preProcessed, training)

```



# Imputation of NA Values

```{r imputation1}
imputeCol <- function(col) {  # impute numeric mean & character mode
	if (class(col)=="numeric") {
		replace(col, which(is.na(col)), mean(col, na.rm=TRUE))
	}
	else {
		replace(col, which(is.na(col)), modeCalc(col))
	}
}

imputeNaN <- function(col) {
	replace(col, which(is.nan(col)), mean(col, na.rm=TRUE))
}

modeCalc <- function(x) {  # mode function
  ux <- unique(x)
  return(ux[which.max(tabulate(match(x, ux)))])
}

checkNA <- function(features) {  # checks NAs
	apply(features, 2, function(x) { length(which(is.na(x))) })
}

hotDeckImpute <- function(col) {  # impute via random sampling from each label class
	set.seed(1)
	if (length(col[which(!is.na(col))])!=0) {
		replace(col, which(is.na(col)), 
			col[which(!is.na(col))][sample(seq(length(which(!is.na(col)))), 1)])
	}
	else {
		col %<>% as.data.frame %>% ungroup %>% unlist %>% as.numeric
		replace(col, which(is.na(col)), 
			col[which(!is.na(col))][sample(seq(length(which(!is.na(col)))), 1)])
	}
}
```



```{r imputation2}
# hot deck imputation 
# some SOFA scores were NAs within one or both labels, prohibiting group_by, fixed by exception handling
featuresHD <- features %>% 
	group_by(label) %>%
	mutate_each(., funs(hotDeckImpute)) %>% 
	ungroup()
featuresHD %>% checkNA
```

```{r}

# mean and mode imputation
# subject IDs are unusable here, use icustay_id instead
rownames(features) <- features$icustay_id
features %<>% 
	group_by(label) %>% 
	mutate_each(., funs(imputeCol)) %>% 
	ungroup() %>%
	mutate_each(., funs(imputeNaN)) %>%
    ## create valid level names
	mutate_each(funs(make.names), matches("descr|gender")) %>%
	mutate_each(funs(as.factor), matches("descr|gender|icustay_first|icustay_intime|label")) %>%
	dplyr::select(-subject_id, -icustay_id) %>%
	filter(icustay_admit_age < 200)

features %>% checkNA


str(features)		 
```

# Descriptives 

```{r}
numSum <- function(featureSet) {
	featureSet %>% 
		group_by(label) %>%
		summarise_each(funs(min(.), max(.), mean(.), median(.), sd(.)), 
					   matches("mean|age|weig|sapsi|sofa"))
}

catSum <- function(featureSet, cat) {
	return(featureSet %>% group_by_(cat, "label") %>% 
		   	summarise(count=n()) %>% ungroup %>% 
		   	group_by_("label") %>% mutate(perc=count/sum(count)))
}

loopDescrPlot <- function(df) {
	ggplot(df, aes_string(colnames(df)[1], "perc")) +
		geom_point(aes(color=as.factor(label), size=perc)) +
		ylab("Percent") + 
		xlab("Trait") + 
		coord_cartesian(xlim=c(0, 1)) +
		ggtitle(colnames(df)[1]) +
		scale_color_manual(values = c("red", "blue")) +
		theme_bw() +
		coord_flip()
}

catVars <- as.list(colnames(features)[c(2, 5:14)])
catStats <- lapply(catVars, catSum, featureSet=features)
catPlots <- lapply(catStats[c(1:7, 10:11)], loopDescrPlot)
catPlots

numSum(features)
numSum(featuresHD)
```

# Center and Scale

```{r}
centerScale <- function(dtm) {
	toScale <- dtm %>% dplyr::select(., matches("mean|weig|sapsi|sofa")) %>% scale()
	dtm %<>% dplyr::select(., -matches("mean|age|weig|sapsi|sofa")) %>% 
		cbind(., toScale)
	return(dtm)
}

features %<>% centerScale
```

# Training and Testing Sets

```{r}
createTrainTest <- function(featureSet) {
	trainIndex <- createDataPartition(featureSet$label, times=1, p=0.7, list=FALSE)
	return(list(featureSet[trainIndex, ], featureSet[-trainIndex, ]))
}

# training [[1]], testing [[2]]
trainTest <- features %>% createTrainTest
```

# Model Creation

```{r}
createModels <- function(trainSet) {
	nlpLogModel <- glm(label ~ ., family=binomial, data=trainSet)
	nlpNaiveBayes <- naiveBayes(as.factor(label) ~ ., data=trainSet, laplace=0)
	nlpInfoModel <- rpart(label ~ ., trainSet, method="class", 
						  parms=list(split="information"))
	nlpGiniModel <- rpart(label ~ ., trainSet, method="class", 
						  parms=list(split="gini"))
	nlpRandomForest <- randomForest(label ~ ., trainSet)
	return(list(nlpLogModel, nlpNaiveBayes, nlpInfoModel, nlpGiniModel))
}
```

# Cross Validation

```{r}
crossValidate <- function(trainSet) {
	tc <- trainControl("cv", 10, savePred=TRUE, classProbs=TRUE, summaryFunction=twoClassSummary)
# 					   classProbs=TRUE, summaryFunction=twoClassSummary)  
	
	logModel <- train(label ~ ., data=trainSet, trControl=tc,  
						 method="glm", metric="ROC")  
	
	nbGrid <- expand.grid(.fL=0, .usekernel=F)
	naiveBayes <- train(label ~ ., data=trainSet, trControl=tc, 
						method="nb", tuneGrid=nbGrid, metric="ROC")  
	
	infoGrid <- expand.grid(.C=0.25)
	infoModel <- train(label ~ ., data=trainSet, trControl=tc,
					   method="J48", tuneGrid=infoGrid, metric="ROC") 
	
	rpartGrid <- expand.grid(.cp=0.2)
	giniModel <- train(label ~ ., data=trainSet, trControl=tc, 
					   method="rpart", tuneGrid=rpartGrid, metric="ROC") 
	
	rfGrid <- expand.grid(.mtry=c(1, 9))
	rfModel <- train(label ~ ., data=trainSet, trControl=tc,
					 method="rf", tuneGrid=rfGrid, metric="ROC")
	
	return(list(logModel, infoModel, giniModel, rfModel))
}

test <- crossValidate(trainTest[[1]])
```

# Test Predictions

```{r}
testResults <- function(models, test, label) {
	
	logitPred <- predict(models[[1]], test, link="class") %>% as.data.frame 
	logitPred <- ifelse(logitPred>0, 1, 0) %>% as.numeric
	logitConMat <- table(logitPred, label) %>% reportStat
	
	nbPred <- predict(models[[2]], test, type="raw") %>% 
		as.data.frame %>% mutate(predLabel=ifelse(`0`>`1`, 0, 1)) %>% 
		dplyr::select(predLabel) %>% unlist %>% as.numeric
	nbConMat <- table(nbPred, label) %>% reportStat
	
	treeInfoPred <- predict(models[[3]], test, link="class") %>% 
		as.data.frame %>% mutate(predLabel=ifelse(`0`>`1`, 0, 1)) %>% 
		dplyr::select(predLabel) %>% unlist %>% as.numeric
	treeInfoConMat <- table(treeInfoPred, label) %>% reportStat
	
	treeGiniPred <- predict(models[[4]], test, link="class") %>% 
		as.data.frame %>% mutate(predLabel=ifelse(`0`>`1`, 0, 1)) %>% 
		dplyr::select(predLabel) %>% unlist %>% as.numeric
	treeGiniConMat <- table(treeGiniPred, label) %>% reportStat
	return(list(logistic=logitConMat, naiveBayes=nbConMat, 
				infoTree=treeInfoConMat, giniTree=treeGiniConMat))
}
```


# SIRS Classifier

```{r}
SIRS <- function(temp, hr, rr, wbc) {
    count <- 0
    if (temp > 38 | temp < 36) count + 1
    if (hr > 90) count + 1
    if (rr > 20) count + 1
    if (wbc > 12000) count + 1
    ifelse(count >= 2, 1, 0)
}


```

